gratia: An R package for exploring generalized additive models

### 2024-11-21 paper.Rmd
address @dill's comments from #296

A GAM has the form
\begin{align*}
y_i &\sim    \mathcal{D}(\mu_i, \phi) \\
g(\mu_i) &=  \mathbf{A}_i\boldsymbol{\gamma} + \sum_{j=1} f_j(x_{ji})
\end{align*}
where observations $y_i$ are assumed to be conditionally distributed $\mathcal{D}$ with expectation $\mathbb{E}(y_i) = \mu_i$ and dispersion parameter $\phi$. The expectation of $y_i$ is given by a linear predictor of strictly parametric terms, whose model matrix is $\mathbf{A}_i$ with parameters $\boldsymbol{\gamma}$, plus a sum of smooth functions of $j = 1, \dots, J$ covariates $f_j()$. $g()$ is a link function mapping values on the linear predictor to the scale of the response.

The smooth functions $f_j$ are represented in the GAM using penalised splines, which are themselves formed as weighted sums of basis functions, $b_k()$, [@De-Boor2001-vg] e.g.
$$
f_j(x_{ij}) = \sum_{k=1}^{K} \beta_{jk} b_{jk}(x_{ij})
$$
for a univariate spline. The weights, $\beta_k$, are model coefficients to be estimated alongside $\boldsymbol{\gamma}$. To avoid overfitting, estimates $\hat{\beta}_{jk}$ and $\hat{\boldsymbol{\gamma}}$ are sought to minimise the penalised log-likelihood of the model
$$
\ell_p(\boldsymbol{\beta}) = \ell(\boldsymbol{\beta}) - \frac{1}{2\phi} \sum_{j} \lambda_{j} \boldsymbol{\beta}^{\mathsf{T}}_j \mathbf{S}_j \boldsymbol{\beta}_j
$$
where $\ell$ and $\ell_p$ are the log likelihood and penalized log likelihood, respectively, of the data at the parameter estimates, $\mathbf{S}_j$ are penalty matrices and $\lambda_{j}$ are smoothing parameters associated with each smooth. Note that $\boldsymbol{\beta}$ now contains the coefficients $\boldsymbol{\gamma}$ and $\beta_{jk}$. $\boldsymbol{\beta}^{\mathsf{T}}_j \mathbf{S}_j \boldsymbol{\beta}_j$ measures the wiggliness of $f_j$, which, with the default penalty, is the integrated squared second derivative of $f_j$. The smoothing parameters, $\boldsymbol{\lambda}$, control the trade-off between fit to the data and the complexity of the estimated functions.

@@

```{r penalised-spline-basis-and-penalty, echo = FALSE, fig.height = 4, fig.width = 10, fig.cap = "\\label{fig:basis-funs}Basis functions (a) and associated penalty matrix (b) for a penalised, low rank, thin plate regression spline. a) shows the individual basis functions (thin coloured lines), as well as the data (black points) to which the GAM was fitted. The estimated smooth is shown as the thick grey line. b) shows the penalty matrix for the basis shown in a). Note the 9th basis function (labelled 'F9', which is the linear function at the lower left to upper right in a), is not affected by the penalty as it has 0 second derivative everywhere, and hence the resulting penalty for this function is 0.", fig.pos = "t!"}
pkgs <- c("mgcv", "gratia", "ggplot2", "dplyr", "tibble", "patchwork")
res <- vapply(pkgs, library, logical(1), character.only = TRUE,
  logical.return = TRUE)
# simulate data
sim_d <- tibble(x = withr::with_seed(1234, runif(100))) |>
  mutate(
    f = gw_f2(x),
    y = withr::with_seed(1234, rnorm(100, mean = f, sd = 1))
  )
# fit model
sim_m <- gam(y ~ s(x, bs = "tp"), data = sim_d, method = "REML",
  drop.intercept = FALSE)
# basis functions for s(x)
bfs <- basis(sim_m, select = "s(x)")
# penalty S
f_x <- bfs |>
  group_by(x) |>
  summarise(.spline = sum(.value), .groups = "drop")
# recentre on the data
bfs <- bfs |>
  mutate(.value = .value + model_constant(sim_m))
f_x <- f_x |>
  mutate(.spline = .spline + model_constant(sim_m))
# plot basis functions and fitted spline
p_bfs <- bfs |>
  ggplot(aes(x = x, y = .value)) +
  geom_line(aes(colour = .bf)) +
  geom_point(data = sim_d, aes(x = x, y = y)) +
  geom_line(data = f_x, aes(x = x, y = .spline),
    linewidth = 1.25, alpha = 0.5) +
  guides(colour = "none") +
  labs(x = "x", y = "y", title = "s(x)")
# plot the penalty
S <- penalty(sim_m, select = "s(x)")
p_S <- S |>
  draw()
p_bfs + p_S + plot_layout(ncol = 2, nrow = 1) +
  plot_annotation(tag_levels = "a", tag_suffix = ")")
```

@@

Taking a Bayesian approach to smoothing with penalized splines [@Kimeldorf1970-cn; @Wahba1983-mi; @Silverman1985-kw; @Wahba1985-bw; see @Miller2019-nf for a summary], it can be shown that GAMs fitted by *mgcv* are an empirical Bayesian model with an improper multivariate normal prior on the basis function coeficients. Samples from the posterior distribution of these models can be used to estimate the uncertainty in quantities derived from a GAM. This can be invaluable in applied research, where, for example, a quantity of interest may arise as an operation on predictions from the model. *gratia* provides functions for sampling from the posterior distribution of estimated smooths and from the model as a whole, where sampling can include the uncertainty in the estimated coefficients (`fitted_samples()`), the sampling uncertainty of the response (`predicted_samples()`), or both (`posterior_samples()`). By default, a Gaussian approximation to the posterior distribution is used, but a simple Metropolis Hasting sampler can be substituted (using `mgcv::gam.mh()`), which has better performance when the posterior is not well approximated by a Gaussian approximation.

@@

Several R packages provide similar functionality to that of *gratia*. Notably, the *mgcViz* package [@Fasiolo2020-nz] provides sophisticated capabilities for plotting estimated smooths for models fitted by *mgcv* and *qgam* [@], and model diagnostics plots. A particular feature of *mgcViz* is that it was designed to be scalable, easily handling models fitted to data with millions of observations, a use case not currently well handled by *gratia*. *tidymv* [@tidymv-2023] and its succesor, *tidygam* [@tidygam-2023], provide plots of estimated smooths and model predictions respectively, as well as differences of smooths, but the focus is on univariate smooths, in constrast to *gratia*'s ability to plot multivariate and specialist smooths (e.g. soap film smooths). Like *gratia*, the *itsadug* package [@itsadug-2022] is motivated to make working with estimated GAMs and related models easier. *itsadug* uses base graphics for plotting, but in place of partial effects plots, produces plots of adjusted predictions, and while useful for a broad range of models, is especially focused on GAMs fitted to longitudinal data with random effects.

@@

A simple GAM for these data is to model the response (`chl`) with a spatial smooth of latitude (`lat`) and longitude (`lon`) as covariates. Here, I use a spline on the sphere (SOS) smoother built using a Duchon spline with second order derivative penalty [@Duchon1977-jr]. Additional terms included in the linear predictor are a smooth of the day of year of sample collection (`jul.day`) and a smooth of ocean depth (`bath`). The response is assumed to be conditionally distributed Tweedie, with the power parameter ($p$) of the distribution estimated during fitting. Model coefficients and smoothing parameters are estimated using restricted maximum likelihood [@Wood2011-kn]
```{r fit-chl-model, cache = TRUE}
ctrl <- gam.control(nthreads = 10)
m1 <- gam(
  chl ~ s(lat, lon, bs = "sos", m = -1, k = 150) +
    s(jul.day, bs = "cr", k = 20) +
    s(bath, k = 10),
  data = chl, method = "REML", control = ctrl, family = tw()
)
```

@@

which show significant heteroscedasticity and departure from the condtional distribution of the response given the model (Figure \ref{fig:m1-appraise}).

@@

The problems with the model aparent in the diagnostics plots are probably due to important controls on chlorophyll *a* missing from the covariates available in the example data. However, the original model assumed constant values for the dispersion, $\phi$, and the power parameter $p$, which may be too inflexible given the absence of important effects in the model.

@@

*gratia* can handle distributional GAMs fitted with *mgcv*, and also those fitted using *GJRM*'s `gamlss()`. Below, the estimated smooths from `m2` are plotted using `draw()`
```{r m2-draw, fig.show = "hide", dependson = "fit-chl-lsslmodel", cache = TRUE, fig.pos = "t!", out.width = ".8\\linewidth"}
# Define the coordinate system to use for plotting on a map
crs <- "+proj=ortho +lat_0=20 +lon_0=-40"
draw(m2, crs = crs, default_crs = 4326, dist = 0.05, rug = FALSE)
```
The plots produced by `draw()` from a fitted model are known as *partial effect plots* (Figure \ref{fig:m2-draw}), which show the component contributions, on the link scale, of each model term to the linear predictor. The y axis on these plots is typically centred around 0 due to most smooths having a sum-to-zero identifiability constraint applied to them. This constraint is what allows the model to include multiple smooths and remain identifiable. These plots allow you to read off the contributions of each smooth to the fitted response (on the link scale); they show link-scale *predictions* of the response for each smooth, conditional upon all other terms in the model, including any parametric effects and the intercept, having zero contribution. In the paralance of the *marginaleffects* package [@Arel-BundockGreiferHeiss:2024], these plots show *adjusted predictions*, just where the adjustment includes setting the contribution of *all* other model terms to the predicted value to zero. For *partial derivatives* (what *marginaleffects* would call a *marginal effect* or slope), *gratia* provides the `derivatives()`.

@@

If the provided plots are insufficient for users' needs, lower-level functionality is provided by *gratia* to facilitate bespoke plotting with *ggplot2*. For example, to evaluate the SOS smooth at a grid (50x50) of values over the range of the covariates, we use `smooth_estimates()` and add a Bayesian credible interval with `add_confint()`:
```{r sm-est-m1, results = "hide", dependson = "fit-chl-lsslmodel", cache = TRUE}
smooth_estimates(m2, select = "s(lat,lon)", n = 50) |>
  add_confint()
```
This returns a data frame of the requested values that is ammenable for plotted with `ggplot()`.

@@

Perhaps we are interested in the average expected chlorophyll *a* between 40--50 degrees N and 40--50 degrees W. An estimate for this value can be computed directly from the fitted model as follows. First create a slice through the data for the spatial locations were are interested in using the `data_slice()` function, which ensures that `ds` contains everything we need to predict from the fitted model
```{r avg-chl-ds, dependson = "fit-chl-lsslmodel", cache = TRUE}
ds <- data_slice(m2,
  lat = evenly(lat, lower = 40, upper = 50, by = 0.5),
  lon = evenly(lon, lower = -50, upper = -40, by = 0.5))
```
Next, `fitted_values()` returns the predicted values at the specified locations. I only include the spatial effects, excluding the effects of ocean depth and day of year, and ignore terms in the other linear predictors as they do not affect the expected chlorophyll *a*:
```{r avg-chl-fv, dependson = c("fit-chl-lsslmodel", "avg-chl-ds"), cache = TRUE}
use <- c("(Intercpt)", "s(lat,lon)")
fv <- fitted_values(m2, data = ds, terms = use) # predict
```
Finally, I summarise the predictions for the location parameter to yield the average of the predicted values
```{r avg-chl-sumamrise, dependson = "avg-chl-fv", cache = TRUE}
fv |>
  filter(.parameter == "location") |>
  summarise(chl_a = mean(.fitted))
```
While this is an acceptable answer to the question, it lacks an uncertainty estimate. This is where posterior sampling is useful. With a small modification of the above code and a little data wrangling, we can produce an uncertainty estimate, using `fitted_samples()` to generate posterior draws of the expected chlorophyll *a*:
```{r post-sampling-1, warning = FALSE, dependson = c("fit-chl-lsslmodel", "avg-chl-ds"), cache = TRUE}
fs <- fitted_samples(m2,     # model
  data = ds,                 # values of covariates to predict at
  terms = use,               # which terms to include in predictions
  n = 10000,                 # number of posterior draws
  method = "gaussian",       # Gaussian approximation to the posterior
  unconditional = TRUE,      # incl uncertainty for estimating lambda
  n_cores = 4,               # how many CPU cores to compute MVN samples
  seed = 342)                # set the random number seed, used internally
```
The posterior draws can then be summarised as before, except now the average chlorophyll *a* is calculated separately for each posterior draw (`.draw`)
```{r summarise-posterior, dependson = "post-sampling-1", cache = TRUE}
fs |>                                 # take the posterior draws
  group_by(.draw) |>                  # group them by `.draw`
  summarise(chl_a = mean(.fitted)) |> # compute mean of fitted chl a
  ggdist::median_qi()                 # summarise posterior
```
The posterior distribution of average chlorophyll *a* is summarised using `median_qi()` from the *ggdist* package [@Kay2024-rv; @Kay2024-uj]. While we could use the base R function `quantile()` to compute the interval, the use of `median_qi()` illustrates how *gratia* tries to interact with other packages.

@@

*gratia* provides a range of functionality to make working with estimated GAMs easier for users. It is designed to take some of the pain out of working with models, simplifying plotting of smooths and related features (differences, derivatives) and exposing the powerful machinary of the *mgcv* package without the need for a deep understanding of GAMs, splines, and the inner structure of *mgcv*'s model objects. As such, it provides a useful addition for anyone wanting to use GAMs in their data analyses without requiring them to be a GAM expert.

>>>>

A GAM has the form
\begin{align*}
y_i &\sim    \mathcal{D}(\mu_i, \phi) \\
g(\mu_i) &=  \mathbf{A}_i\boldsymbol{\gamma} + f_1(x_{1i}) + f_2(x_{2i}) + f_3(x_{3i}, x_{4i}) + \cdots
\end{align*}
where observations $y_i$ are assumed to be conditionally distributed $\mathcal{D}$ with expectation $\mathbb{E}(y_i) = \mu_i$ and dispersion parameter $\phi$. The expectation of $y_i$ is given by a linear predictor of strictly parametric terms, whose model matrix is $\mathbf{A}_i$ with parameters $\boldsymbol{\gamma}$, plus a sum of $j = 1, \dots, J$ smooth functions of covariates $f_j()$. $g()$ is a link function mapping values on the linear predictor to the scale of the response.

The smooth functions $f_j$ are represented in the GAM using penalised splines, which are themselves formed as weighted sums of basis functions, $b_k()$, [@De-Boor2001-vg] e.g.
$$
f_j(x_{ij}) = \sum_{k=1}^{K} \beta_{jk} b_{jk}(x_{ij})
$$
for a univariate spline. The weights, $\beta_k$, are model coefficients to be estimated alongside $\boldsymbol{\gamma}$. To avoid overfitting, estimates $\hat{\beta}_{jk}$ and $\hat{\boldsymbol{\gamma}}$ are sought to minimise the penalised log-likelihood of the model
$$
\ell_p(\boldsymbol{\beta}) = \ell(\boldsymbol{\beta}) - \frac{1}{2\phi} \sum_{j} \lambda_{j} \boldsymbol{\beta}^{\mathsf{T}}_j \mathbf{S}_j \boldsymbol{\beta}_j
$$
where $\ell$ and $\ell_p$ are the log likelihood and penalized log likelihood, respectively, of the data at the parameter estimates, $\mathbf{S}_j$ are penalty matrices and $\lambda_{j}$ are smoothing parameters associated with each smooth. Note that $\boldsymbol{\beta}$ now contains the coefficients $\boldsymbol{\gamma}$ and $\beta_{jk}$. $\boldsymbol{\beta}^{\mathsf{T}}_j \mathbf{S}_j \boldsymbol{\beta}_j$ measures the wiggliness of $f_j$, which, with the default basis-penalty combination, is the integrated squared second derivative of $f_j$. The smoothing parameters, $\boldsymbol{\lambda}$, control the trade-off between fit to the data and the complexity of the estimated functions.

@@

```{r penalised-spline-basis-and-penalty, echo = FALSE, fig.height = 4, fig.width = 10, fig.cap = "\\label{fig:basis-funs}Basis functions (a) and associated penalty matrix (b) for a penalised, low rank, thin plate regression spline. a) shows the individual basis functions (thin coloured lines) multiplied by their respective model coefficients, as well as the data (black points) to which the GAM was fitted. The estimated smooth is shown as the thick grey line. b) shows the penalty matrix for the basis shown in a). Note the 9th basis function (labelled 'F9', which is the linear function at the lower left to upper right in a), is not affected by the penalty as it has 0 second derivative everywhere, and hence the resulting penalty for this function is 0.", fig.pos = "t!"}
pkgs <- c("mgcv", "gratia", "ggplot2", "dplyr", "tibble", "patchwork")
res <- vapply(pkgs, library, logical(1), character.only = TRUE,
  logical.return = TRUE)
# simulate data
sim_d <- tibble(x = withr::with_seed(1234, runif(100))) |>
  mutate(
    f = gw_f2(x),
    y = withr::with_seed(1234, rnorm(100, mean = f, sd = 1))
  )
# fit model
sim_m <- gam(y ~ s(x, bs = "tp"), data = sim_d, method = "REML",
  drop.intercept = FALSE)
# basis functions for s(x)
bfs <- basis(sim_m, select = "s(x)")
# penalty S
f_x <- bfs |>
  group_by(x) |>
  summarise(.spline = sum(.value), .groups = "drop")
# recentre on the data
bfs <- bfs |>
  mutate(.value = .value + model_constant(sim_m))
f_x <- f_x |>
  mutate(.spline = .spline + model_constant(sim_m))
# plot basis functions and fitted spline
p_bfs <- bfs |>
  ggplot(aes(x = x, y = .value)) +
  geom_line(aes(colour = .bf)) +
  geom_point(data = sim_d, aes(x = x, y = y)) +
  geom_line(data = f_x, aes(x = x, y = .spline),
    linewidth = 1.25, alpha = 0.5) +
  guides(colour = "none") +
  labs(x = "x", y = "y", title = "s(x)")
# plot the penalty
S <- penalty(sim_m, select = "s(x)")
p_S <- S |>
  draw()
p_bfs + p_S + plot_layout(ncol = 2, nrow = 1) +
  plot_annotation(tag_levels = "a", tag_suffix = ")")
```

@@

Taking a Bayesian approach to smoothing with penalized splines [@Kimeldorf1970-cn; @Wahba1983-mi; @Silverman1985-kw; @Wahba1985-bw; see @Miller2019-nf for a summary], it can be shown that GAMs fitted by *mgcv* are an empirical Bayesian model with a (usually improper) multivariate normal prior on the basis function coefficients. Samples from the posterior distribution of these models can be used to estimate the uncertainty in quantities derived from a GAM. This can be invaluable in applied research, where, for example, a quantity of interest may arise as an operation on predictions from the model. *gratia* provides functions for sampling from the posterior distribution of estimated smooths and from the model as a whole, where sampling can include the uncertainty in the estimated coefficients (`fitted_samples()`), the sampling uncertainty of the response (`predicted_samples()`), or both (`posterior_samples()`). By default, a Gaussian approximation to the posterior distribution is used, but a simple Metropolis Hastings sampler can be substituted (using `mgcv::gam.mh()`), which has better performance when the posterior is not well approximated by a Gaussian approximation.

@@

Several R packages provide similar functionality to that of *gratia*. Notably, the *mgcViz* package [@Fasiolo2020-nz] provides sophisticated capabilities for plotting estimated smooths for models fitted by *mgcv* and *qgam* [@], and model diagnostics plots. A particular feature of *mgcViz* is that it was designed to be scalable, easily handling models fitted to data with millions of observations, a use case not currently well handled by *gratia*. *tidymv* [@tidymv-2023] and its successor, *tidygam* [@tidygam-2023], provide plots of estimated smooths and model predictions respectively, as well as differences of smooths, but the focus is on univariate smooths, in contrast to *gratia*'s ability to plot multivariate and specialist smooths (e.g. soap film smooths). Like *gratia*, the *itsadug* package [@itsadug-2022] is motivated to make working with estimated GAMs and related models easier. *itsadug* uses base graphics for plotting, but in place of partial effects plots, produces plots of adjusted predictions, and while useful for a broad range of models, is especially focused on GAMs fitted to longitudinal data with random effects.

@@

A simple GAM for these data is to model the response (`chl`) with a spatial smooth of latitude (`lat`) and longitude (`lon`) as covariates. Here, I use a spline on the sphere (SOS) smoother built using a Duchon spline with second order derivative penalty [@Duchon1977-jr]. Additional terms included in the linear predictor are a smooth of the day of year of sample collection (`jul.day`) and a smooth of ocean depth (`bath`). The response is assumed to be conditionally Tweedie distributed, with the power parameter ($p$) of the distribution estimated during fitting. Model coefficients and smoothing parameters are estimated using restricted maximum likelihood [@Wood2011-kn]. To use parallel processing in some parts of the model fitting algorithm, the `nthreads` control parameter is set to 10 (set this lower if your machine doesn't have this many physical CPU cores).
```{r fit-chl-model, cache = TRUE}
ctrl <- gam.control(nthreads = 10)
m1 <- gam(
  chl ~ s(lat, lon, bs = "sos", m = -1, k = 150) +
    s(jul.day, bs = "cr", k = 20) +
    s(bath, k = 10),
  data = chl, method = "REML", control = ctrl, family = tw()
)
```

@@

which show significant heteroscedasticity and departure from the conditional distribution of the response given the model, as seen in the increasing spread of the deviance residuals at larger values of the linear predictor in the upper right panel of Figure \ref{fig:m1-appraise}.

@@

The problems with the model apparent in the diagnostics plots are probably due to important controls on chlorophyll *a* missing from the covariates available in the example data. However, the original model assumed constant values for the dispersion, $\phi$, and the power parameter $p$, which may be too inflexible if missing covariates mean that important effects are not included in the model.

@@

*gratia* can handle distributional GAMs fitted with *mgcv*, and also those fitted using *GJRM*'s `gamlss()`. Below, the estimated smooths from `m2` are plotted using `draw()`
```{r m2-draw, fig.show = "hide", dependson = "fit-chl-lsslmodel", cache = TRUE, fig.pos = "t!", out.width = ".8\\linewidth"}
# Define the coordinate system to use for plotting on a map
crs <- "+proj=ortho +lat_0=20 +lon_0=-40"
draw(m2, crs = crs, default_crs = 4326, dist = 0.05, rug = FALSE)
```
The plots produced by `draw()` from a fitted model are known as *partial effect plots* (Figure \ref{fig:m2-draw}), which show the component contributions, on the link scale, of each model term to the linear predictor. The y axis on these plots is typically centred around 0 due to most smooths having a sum-to-zero identifiability constraint applied to them. This constraint is what allows the model to include multiple smooths and remain identifiable. These plots allow you to read off the contributions of each smooth to the fitted response (on the link scale); they show link-scale *predictions* of the response for each smooth, conditional upon all other terms in the model, including any parametric effects and the intercept, having zero contribution. In the parlance of the *marginaleffects* package [@Arel-BundockGreiferHeiss:2024], these plots show *adjusted predictions*, just where the adjustment includes setting the contribution of *all* other model terms to the predicted value to zero. For *partial derivatives* (what *marginaleffects* would call a *marginal effect* or slope), *gratia* provides the `derivatives()`.

@@

If the provided plots are insufficient for users' needs, lower-level functionality is provided by *gratia* to facilitate bespoke plotting with *ggplot2*. For example, to evaluate the SOS smooth at a grid (50x50) of values over the range of the covariates, we use `smooth_estimates()` and add a Bayesian credible interval with `add_confint()`:
```{r sm-est-m1, results = "hide", dependson = "fit-chl-lsslmodel", cache = TRUE}
smooth_estimates(m2, select = "s(lat,lon)", n = 50) |>
  add_confint()
```
This returns a data frame of the requested values that is amenable for plotted with `ggplot()`.

@@

Perhaps we are interested in the average expected chlorophyll *a* between 40--50 degrees N and 40--50 degrees W. An estimate for this value can be computed directly from the fitted model as follows. First create a slice through the data for the spatial locations were are interested in using the `data_slice()` function, which ensures that `ds` contains everything we need to predict from the fitted model
```{r avg-chl-ds, dependson = "fit-chl-lsslmodel", cache = TRUE}
ds <- data_slice(m2,
  lat = evenly(lat, lower = 40, upper = 50, by = 0.5),
  lon = evenly(lon, lower = -50, upper = -40, by = 0.5))
```
Next, `fitted_values()` returns the predicted values at the specified locations. I only include the spatial effects, excluding the effects of ocean depth and day of year, and ignore terms in the other linear predictors as they do not affect the expected chlorophyll *a*:
```{r avg-chl-fv, dependson = c("fit-chl-lsslmodel", "avg-chl-ds"), cache = TRUE}
use <- c("(Intercept)", "s(lat,lon)")
fv <- fitted_values(m2, data = ds, terms = use) # predict
```
Finally, I summarise the predictions for the location parameter to yield the average of the predicted values
```{r avg-chl-summarise, dependson = "avg-chl-fv", cache = TRUE}
fv |>
  filter(.parameter == "location") |>
  summarise(chl_a = mean(.fitted))
```
While this is an acceptable answer to the question, it lacks an uncertainty estimate. This is where posterior sampling is useful. With a small modification of the above code and a little data wrangling, we can produce an uncertainty estimate, using `fitted_samples()` to generate posterior draws of the expected chlorophyll *a*:
```{r post-sampling-1, warning = FALSE, dependson = c("fit-chl-lsslmodel", "avg-chl-ds"), cache = TRUE}
fs <- fitted_samples(m2,     # model
  data = ds,                 # values of covariates to predict at
  terms = use,               # which terms to include in predictions
  n = 10000,                 # number of posterior draws
  method = "gaussian",       # Gaussian approximation to the posterior
  unconditional = TRUE,      # incl uncertainty for estimating lambda
  n_cores = 4,               # how many CPU cores to compute MVN samples
  seed = 342)                # set the random number seed, used internally
```
The posterior draws can then be summarised as before, except now the average chlorophyll *a* is calculated separately for each posterior draw (`.draw`)
```{r summarise-posterior, dependson = "post-sampling-1", cache = TRUE}
fs |>                                 # take the posterior draws
  group_by(.draw) |>                  # group them by `.draw`
  summarise(chl_a = mean(.fitted)) |> # compute mean of fitted chl a
  ggdist::median_qi()                 # summarise posterior
```
The posterior distribution of average chlorophyll *a* is summarised using `median_qi()` from the *ggdist* package [@Kay2024-rv; @Kay2024-uj]. While we could use the base R function `quantile()` to compute the interval, the use of `median_qi()` illustrates how *gratia* tries to interact with other packages.

@@

*gratia* provides a range of functionality to make working with estimated GAMs easier for users. It is designed to take some of the pain out of working with models, simplifying plotting of smooths and related features (differences, derivatives) and exposing the powerful machinery of the *mgcv* package without the need for a deep understanding of GAMs, splines, and the inner structure of *mgcv*'s model objects. As such, it provides a useful addition for anyone wanting to use GAMs in their data analyses without requiring them to be a GAM expert.
